Supplemental material: Ecosystem carbon balance in the Hawaiian Islands under different scenarios of future climate and land use change

### 2021-02-16 Manuscript/HI_Model_ms.Rmd
updated Fire R scripts and manuscript files

## Introduction
Terrestrial ecosystems are a major sink for atmospheric carbon dioxide (CO~2~), removing ~30% of human emissions on an annual basis and reducing the rate of increase in atmospheric CO~2~ [@keenan_terrestrial_2018; @friedlingstein_global_2019]. There is increasing recognition among policymakers that natural and angricultural ecosystems can contribute to climate mitigation, which has given rise to the popularity of “natural carbon solutions” [@cameron_ecosystem_2017]. Defined as conservation and land management efforts aimed at enhancing ecosystem carbon storage [@griscom_natural_2017], natural climate solutions are appealing because they are seen as cost-effective and readily available [@averchenkova_climate_2017; @cameron_ecosystem_2017; @fargione_natural_2018]. However, effective implementation is complicated by the uncertainty surrounding the future direction and magnitude of the land carbon sink, especially at the regional scale. Despite this uncertainty, evidence indicates that both interannual and long-term variability in carbon uptake by land ecosystems is driven primarily by fluctuations in climate, land use, and land cover change [@ahlstrom_dominant_2015; @prestele_current_2017; @friedlingstein_global_2019]. Incorporating the interactive effects of land use and climate into spatially explicit future projections of ecosystem carbon balance could therefore provide a reference point to evaluate the effectiveness of land-based mitigation. Although a complex challenge, the growing number of sub-national jurisdictions that plan to incorporate land-based mitigation strategies into their emissions reduction efforts would benefit from understanding how future land-use and climate-biosphere feedbacks will affect ecosystem carbon balance in their respective regions [@sleeter_effects_2019].

@@

Here we consider four unique scenarios exploring a range of assumptions about future climate, land use, land cover, disturbance, and global CO~2~ emissions. We used a stochastic, spatially explicit simulation model to estimate ecosystem carbon balance for Hawaiʻi's natural and agricultural lands under each of these four scenarios on an annual basis for the period 2010–2100 [@daniel_state-and-transition_2016; @daniel_integrating_2018; @sleeter_effects_2019]. The four unique scenarios represent all combinations of two land-use change pathways (low and high) and two radiative forcing pathways (representative concentration pathway [RCP] 4.5 and 8.5). In addition to these four scenarios, we conducted a separate series of simulations to examine how ecosystem carbon balance estimates vary in response to different levels of a CO~2~ fertilization effect (CFE) on net primary productivity [NPP; @sleeter_effects_2019]. our goals were to estimate changes in Hawaiian ecosystem carbon balance and their uncertainties under a range of plausible future scenarios, quantify the relative impact of major controlling processes such as land use change, disturbance, and climate change, and assess the sensitivity of model estimates to the introduction of a CFE on NPP.

## Methods
We used the Land Use and Carbon Scenario Simulator (LUCAS), an integrated landscape change and carbon gain-loss model, to project changes in ecosystem carbon balance for the seven main Hawaiian Islands under all combinations of two land-use scenarios (low and high) and two radiative forcing scenarios (RCP 4.5 and RCP 8.5). We also developed a separate set of scenarios to test model sensitivity to different levels of a CO~2~ fertilization effect (CFE). The landscape change portion of LUCAS is a state-and-transition model that applies a Monte Carlo approach to track the state type and age of each simulation cell in response to a pre-determined set of transitions [@daniel_state-and-transition_2016]. The carbon gain-loss portion tracks carbon stocks within each simulation cell over time as continuous state variables, along with a pre-defined set of continuous flows specifying rates of change in stock levels over time [@daniel_integrating_2018; @sleeter_effects_2019]. We parameterized the Hawaiʻi LUCAS model to estimate annual changes in carbon stocks and fluxes in response to land use, land use change, wildland fire, and long-term climate variability for the time period 2010-2100.

### *Study area*
The spatial extent of this study was the terrestrial portion of the seven main Hawaiian Islands (figure 1), a total land area of 16,554 km^2^. We subdivided this landscape into a grid of 264,870 simulation cells, each of which was 250 x 250 m in size. Each simulation cell was assigned to one of 210 possible state types based on the unique combination of three moisture zones (dry, mesic, and wet; figure S1), seven islands, and ten discrete land cover classes (figure 1).

@@

### *States and transitions*
We developed two land-use scenarios (low and high) with transition pathways modified from @daniel_state-and-transition_2016. Transitions between state types were pre-defined to represent urbanization, agricultural contraction, agricultural expansion, harvesting of tree plantations, and wildfire. Agriculture, forest, grassland, tree plantation, and shrubland state types each had multiple transition pathways, while the barren state type could only transition to developed (i.e., urbanization). There was no transition pathway out of an urbanized (developed) state. Water and wetland state types remained static throughout the simulation period.

@@

### *Carbon stocks and flows*
 The fate of carbon stocks was tracked for each simulation cell based on a suite of carbon flows (i.e., carbon fluxes) specifying the rates of change in these carbon stocks over time [@daniel_integrating_2018; @sleeter_effects_2019]. We defined carbon stocks as continuous state variables for each simulation cell, including live biomass, standing dead wood, down dead wood, litter, and soil organic carbon. We also included and tracked carbon in atmospheric, aquatic, and harvest product pools to enforce carbon mass balance [@daniel_integrating_2018]. To transfer carbon between stocks, we defined baseline carbon flows as continuous variables resulting from growth, mortality, deadfall, woody decay, litter decomposition, and leaching (which includes runoff). We also defined carbon flows resulting from land use, land use change, and wildfire [@selmants_baseline_2017; @daniel_integrating_2018].

@@

### *CO~2~ fertilization effect*
Increasing atmospheric CO~2~ concentrations stimulate leaf-level photosynthesis, potentially increasing NPP as well [@Walker_integrating_2020]. However, the magnitude and persistence of this effect is highly uncertain, particularly across a range of climatic conditions and over long time spans [@Walker_integrating_2020]. Following @sleeter_effects_2019, we developed a separate set of scenarios designed to test the sensitivity of LUCAS model projections of ecosystem carbon balance to different rates of a CO~2~ fertilization effect (CFE). We incorporated a CFE multiplier for NPP that represented the percent increase in NPP for every 100 ppm increase in atmospheric CO~2~ concentration under the high land use and high radiative forcing (RCP 8.5) scenario. We tested five CFE levels ranging from 5% to 15%, which is within the range of CFEs observed in free air CO~2~ enrichment (FACE) experiments. For all levels, we assumed CFEs reached saturation at an atmospheric CO~2~ concentration of 600 ppm, with no further stimulation of NPP despite a continued increase in CO~2~ concentration to 930 ppm by 2100. This 600ppm threshold generally coincides with the upper limit from FACE experiments and is reached by the year 2060 under RCP 8.5.

### *Scenario simulations and analysis*
Each of the four unique scenarios were run for 90 years at an annual timestep and repeated for 30 Monte Carlo realizations, using initial conditions corresponding to the year 2010. All simulations were performed within the SyncroSim (version 2.2.4) software framework with ST-Sim (version 3.2.13) and SF (version 3.2.10) add-on modules [@daniel_state-and-transition_2016; @daniel_integrating_2018]. Model inputs and outputs were prepared with the R statistical computing platform [@RCore_2019] using the tidyverse [@Wickham_tidyverse_2019], raster [@hijmans_raster_2020], and rsyncrosim [@daniel_rsyncrosim_2020] packages. Carbon stocks and fluxes for the seven main Hawaiian Islands were calculated for each scenario by summing within each Monte Carlo realization on an annual basis and then calculating annual means as well as the annual upper and lower limits of the 30 Monte Carlo realizations. Carbon balance for the seven main Hawaiian Islands was calculated on annual basis for each scenario and Monte Carlo realization as net biome productivity (NBP), which was equal to annual carbon input in the form of NPP minus the annual sum of all carbon losses from terrestrial ecosystems, including heterotrophic respiration (R~h~) from litter and soil, carbon fluxes to the atmosphere triggered by land use and land-use change, wildfire emissions, and aquatic carbon losses through leaching and overland flow. Positive NBP values indicated ecosystems of the seven main Hawaiian Islands were acting as a net sink for atmospheric CO~2~, while negative NBP values indicated that these ecosystems were acting as a net carbon source to the atmosphere [@chapin_reconciling_2006].

## Results

### *Carbon stocks and fluxes*
Terrestrial ecosystems of the seven main Hawaiian Islands stored an estimated 316 Tg of carbon at the beginning of the simulation period in 2010 (figure 2a), with 58% in soil organic matter, 22% in living biomass, and 20% in surface dead organic matter (litter and dead wood; figure 2b). Ecosystems accumulated carbon in all scenarios but at different rates, with trajectories shaped primarily by climate change and to a lesser extent by land-use change. The highest and most consistent projected accumulation of ecosystem carbon occured under the combination of low radiative forcing and low land use change, yielding a ~15% increase in ecosystem carbon to an average of 363 Tg by 2100 (figure 2a). In contrast, high radiative forcing and high land use change resulted in the lowest ecosystem carbon gain, reaching a peak of ~332 Tg in 2063 and a decline to 327 Tg in 2100, resulting in a net increase of only 3% by the end of the simulation period (figure 2a). Ecosystem carbon accumulation was driven exclusively by increasing soil organic carbon across all four scenarios, all other stocks declined over time (figure 2b).

@@

### *Ecosystem carbon balance*
Net biome productivity (NBP) averaged approximately 0.6 Tg C y^-1^ at the start of the simulation period and declined over time in all four scenarios (figure 4). On average, terrestrial ecosystems of the seven main Hawaiian Islands collectively acted as a net carbon sink throughout the simulation period under the RCP 4.5 radiative forcing scenario, but carbon sink strength was ~40% lower in the high land-use scenario compared to the low land-use scenario by the end of the simulation period (figure 4). In contrast, ecosystems of the Hawaiian Islands acted as a net carbon source to the atmosphere toward the latter half of the simulation period under RCP 8.5, with the transition from sink to source occuring 15 years earlier on average in the high land-use scenario than in the low land-use scenario (figure 4). The high land-use scenario under RCP 8.5 represented a ~40% larger net source of carbon to the atmosphere by the year 2100 than the low-land use scenario under the same radiative forcing. Over the entire simulation period, both global emissions reductions and local avoided land conversion resulted in substantial increases in cumulative NBP (figure 5). However, switching from RCP 8.5 to RCP 4.5 increased cumulative NBP in the Hawaiian Islands more than twice as much as reducing emissions from local land-use change and wildfire disturbance (figure 5). Switching from RCP 8.5 to RCP 4.5 under the low land-use scenario yielded the greatest cumulative increase in NBP, resulting in a median gain of 26.5 Tg of carbon over the entire 90-year simulation period.

@@

### *CO~2~ fertilization effect*
Projected estimates of both total ecosystem carbon storage and ecosystem carbon balance were highly sensitive to differing rates of a CFE on plant productivity. Under the high radiative forcing (RCP 8.5) and high land-use scenario, the inclusion of a CFE ranging from 5-15% led to ~33-98 Tg of additional carbon storage in ecosystems by the end of the century, a ~10-30% increase (figure 6a). Compared to the reference scenario (0% CFE), a 5% CFE was sufficient to transform Hawaiian Island ecosystems from a net carbon source to the atmosphere for the latter half of the 21st century (figure 4b) to a net carbon sink for the entire simulation period (figure 6b), completely offsetting all other carbon losses induced by high radiative forcing and high land use. Net carbon sink strength was further enhanced at higher CFE rates, with NBP increasing by an average of 0.07 Tg C y^-1^ for each 1% increase in CFE (figure 6b). When compared to other scenarios, applying a 5% CFE to the high radiative forcing and high land-use scenario resulted in a mean annual NBP of 0.46 ± 0.3 Tg C y^-1^, roughly equivalent to mean annual NBP in the low radiative forcing and low land-use scenario with no CFE (0.52 ± 0.12). A 15% CFE applied to the high radiative forcing and high land-use scenario resulted in a mean annual NBP of 1.18 ± 0.29 Tg C y^-1^, more than double that of the low radiative forcing and low land-use scenario with no CFE.

@@

## Discussion

We estimated that terrestrial ecosystems of the Hawaiian Islands have been a consistent net sink for atmospheric carbon over the last decade (figure 4). For the time period 2011-2019, net biome productivity (NBP) averaged 0.64 TgC y^-1^ and ranged from 0.46 to 0.88 TgC y^-1^ across all scenarios. Based on this mean annual NBP estimate, Hawaiian terrestrial ecosystems offset approximately 13% of 2015 statewide CO~2~ emissions from energy production and transportation (5.04 TgC), the State of Hawaiʻi's largest source of greenhouse gas emissions [@hawaii_inventory_2019]. Future projections indicate Hawaiian terrestrial ecosystems will continue to be a net sink for atmospheric carbon if global CO~2~ emissions peak around 2040 and then decline (RCP 4.5), and that carbon sink strength can be further enhanced by reducing the intensity and extent of future land use change. If, however, global CO~2~ emissions continue to rise throughout the 21st century (RCP 8.5), our projections indicate Hawaiian ecosystems will transition from a net sink to a net source of CO~2~ to the atmosphere, with high levels of land-use change accelerating this transition and exacerbating net carbon loss. Our model results also indicate that projections of ecosystem carbon balance are highly sensitive to the introduction of a CFE. Even a 5% increase in NPP for every 100ppm increase in atmospheric CO~2~ was sufficient to completely offset all other carbon losses induced by the high radiative forcing and high land use scenario, maintaining Hawaiian Island ecosystems as a net carbon sink for the entire simulation period instead of transitioning to a net carbon source by mid-century. Reconciling the high uncertainty surrounding the response of net photosynthesis to rising atmospheric CO~2~ is essential to more realistically constrain simulation model projections.

### *Drivers of ecosystem carbon balance*

### *Comparison to other studies*
There are very few estimates of contemporary ecosystem carbon balance for the main Hawaiian Islands, and even fewer model projections of future ecosystem carbon balance.  Our mean annual NBP estimate of 0.64 TgC y^-1^ for the period 2011-2019 agrees well with a recent State of Hawaiʻi Greenhouse Gas Inventory report, which estimated an annual net carbon sink of 0.66 TgC in 2015 from agriculture, forestry, and other land uses [@hawaii_inventory_2019].

### *Limitations of this study*

## Conclusion

>>>>

## 1. Introduction
Terrestrial ecosystems are a major sink for atmospheric carbon dioxide (CO~2~), removing ~30% of human emissions on an annual basis and reducing the rate of increase in atmospheric CO~2~ [@keenan_terrestrial_2018; @friedlingstein_global_2019]. There is increasing recognition among policymakers that natural and angricultural ecosystems can contribute to climate mitigation, which has given rise to the popularity of “natural carbon solutions” [@cameron_ecosystem_2017]. Defined as conservation and land management efforts aimed at enhancing ecosystem carbon storage [@griscom_natural_2017], natural climate solutions are appealing because they are seen as cost-effective and readily available [@averchenkova_climate_2017; @cameron_ecosystem_2017; @fargione_natural_2018]. However, effective implementation is complicated by the uncertainty surrounding the future direction and magnitude of the land carbon sink, especially at the regional scale. Despite this uncertainty, evidence indicates that both interannual and long-term variability in carbon uptake by land ecosystems is driven primarily by fluctuations in climate, land use, and land cover change [@ahlstrom_dominant_2015; @prestele_current_2017; @friedlingstein_global_2019]. Incorporating the interactive effects of land use and climate into spatially explicit future projections of ecosystem carbon balance could therefore provide a reference point to evaluate the effectiveness of land-based mitigation. Although a complex challenge, the growing number of sub-national jurisdictions that plan to incorporate land-based mitigation strategies into their emissions reduction efforts would benefit from understanding how future land-use and climate-biosphere feedbacks will affect ecosystem carbon balance in their respective regions [@sleeter_effects_2019].

@@

We used a stochastic, spatially explicit simulation model to estimate ecosystem carbon balance for Hawaiʻi's natural and agricultural lands on an annual basis for the period 2010–2100 under a range of assumptions about future climate, land use, land cover, disturbance, and global CO~2~ emissions  [@daniel_state-and-transition_2016; @daniel_integrating_2018; @sleeter_effects_2019]. We explored four unique scenarios that represent all combinations of two land-use change pathways (low and high) and two radiative forcing pathways (representative concentration pathway [RCP] 4.5 and 8.5). In addition to these four scenarios, we conducted a separate series of simulations to examine how ecosystem carbon balance estimates vary in response to different levels of a CO~2~ fertilization effect (CFE) on net primary productivity [NPP; @sleeter_effects_2019]. our goals were to estimate changes in Hawaiian ecosystem carbon balance and their uncertainties under a range of plausible future scenarios, quantify the relative impact of major controlling processes such as land use change, disturbance, and climate change, and assess the sensitivity of model estimates to the introduction of a CFE on NPP.

## 2. Methods
We used the Land Use and Carbon Scenario Simulator (LUCAS), an integrated landscape change and carbon gain-loss model, to project changes in ecosystem carbon balance for the seven main Hawaiian Islands under all combinations of two land-use scenarios (low and high) and two radiative forcing scenarios (RCP 4.5 and RCP 8.5). We also developed a separate set of scenarios to test model sensitivity to different levels of a CO~2~ fertilization effect (CFE). The landscape change portion of LUCAS is a state-and-transition model that applies a Monte Carlo approach to track the state type and age of each simulation cell in response to a pre-determined set of transitions [@daniel_state-and-transition_2016]. The carbon gain-loss portion tracks carbon stocks within each simulation cell over time as continuous state variables, along with a pre-defined set of continuous flows specifying rates of change in stock levels over time [@daniel_integrating_2018; @sleeter_effects_2019]. We parameterized the Hawaiʻi LUCAS model to estimate annual changes in carbon stocks and fluxes in response to land use, land use change, wildland fire, and long-term climate variability for the time period 2010-2100.

### *2.1 Study area*
The spatial extent of this study was the terrestrial portion of the seven main Hawaiian Islands (figure 1), a total land area of 16,554 km^2^. We subdivided this landscape into a grid of 264,870 simulation cells, each of which was 250 x 250 m in size. Each simulation cell was assigned to one of 210 possible state types based on the unique combination of three moisture zones (dry, mesic, and wet; figure S1), seven islands, and ten discrete land cover classes (figure 1).

@@

### *2.2 States and transitions*
We developed two land-use scenarios (low and high) with transition pathways modified from @daniel_state-and-transition_2016. Transitions between state types were pre-defined to represent urbanization, agricultural contraction, agricultural expansion, harvesting of tree plantations, and wildfire. Agriculture, forest, grassland, tree plantation, and shrubland state types each had multiple transition pathways, while the barren state type could only transition to developed (i.e., urbanization). There was no transition pathway out of an urbanized (developed) state. Water and wetland state types remained static throughout the simulation period.

@@

### *2.3 Carbon stocks and flows*
 The fate of carbon stocks was tracked for each simulation cell based on a suite of carbon flows (i.e., carbon fluxes) specifying the rates of change in these carbon stocks over time [@daniel_integrating_2018; @sleeter_effects_2019]. We defined carbon stocks as continuous state variables for each simulation cell, including live biomass, standing dead wood, down dead wood, litter, and soil organic carbon. We also included and tracked carbon in atmospheric, aquatic, and harvest product pools to enforce carbon mass balance [@daniel_integrating_2018]. To transfer carbon between stocks, we defined baseline carbon flows as continuous variables resulting from growth, mortality, deadfall, woody decay, litter decomposition, and leaching (which includes runoff). We also defined carbon flows resulting from land use, land use change, and wildfire [@selmants_baseline_2017; @daniel_integrating_2018].

@@

### *2.4 CO~2~ fertilization effect*
Increasing atmospheric CO~2~ concentrations stimulate leaf-level photosynthesis, potentially increasing NPP as well [@Walker_integrating_2020]. However, the magnitude and persistence of this effect is highly uncertain, particularly across a range of climatic conditions and over long time spans [@Walker_integrating_2020]. Following @sleeter_effects_2019, we developed a separate set of scenarios designed to test the sensitivity of LUCAS model projections of ecosystem carbon balance to different rates of a CO~2~ fertilization effect (CFE). We incorporated a CFE multiplier for NPP that represented the percent increase in NPP for every 100 ppm increase in atmospheric CO~2~ concentration under the high land use and high radiative forcing (RCP 8.5) scenario. We tested five CFE levels ranging from 5% to 15%, which is within the range of CFEs observed in free air CO~2~ enrichment (FACE) experiments. For all levels, we assumed CFEs reached saturation at an atmospheric CO~2~ concentration of 600 ppm, with no further stimulation of NPP despite a continued increase in CO~2~ concentration to 930 ppm by 2100. This 600ppm threshold generally coincides with the upper limit from FACE experiments and is reached by the year 2060 under RCP 8.5.

### *2.5 Scenario simulations and analysis*
Each of the four unique scenarios were run for 90 years at an annual timestep and repeated for 30 Monte Carlo realizations, using initial conditions corresponding to the year 2010. All simulations were performed within the SyncroSim (version 2.2.4) software framework with ST-Sim (version 3.2.13) and SF (version 3.2.10) add-on modules [@daniel_state-and-transition_2016; @daniel_integrating_2018]. Model inputs and outputs were prepared with the R statistical computing platform [@RCore_2019] using the tidyverse [@Wickham_tidyverse_2019], raster [@hijmans_raster_2020], and rsyncrosim [@daniel_rsyncrosim_2020] packages. Carbon stocks and fluxes for the seven main Hawaiian Islands were calculated for each scenario by summing within each Monte Carlo realization on an annual basis and then calculating annual means as well as the annual upper and lower limits of the 30 Monte Carlo realizations. Carbon balance for the seven main Hawaiian Islands was calculated on annual basis for each scenario and Monte Carlo realization as net biome productivity (NBP), which was equal to annual carbon input in the form of NPP minus the annual sum of all carbon losses from terrestrial ecosystems, including heterotrophic respiration (R~h~) from litter and soil, carbon fluxes to the atmosphere triggered by land use and land-use change, wildfire emissions, and aquatic carbon losses through leaching and overland flow. Positive NBP values indicated ecosystems of the seven main Hawaiian Islands were acting as a net sink for atmospheric CO~2~, while negative NBP values indicated that these ecosystems were acting as a net carbon source to the atmosphere [@chapin_reconciling_2006].

## 3. Results

### *3.1 Carbon stocks and fluxes*
Terrestrial ecosystems of the seven main Hawaiian Islands stored an estimated 316 Tg of carbon at the beginning of the simulation period in 2010 (figure 2a), with 58% in soil organic matter, 22% in living biomass, and 20% in surface dead organic matter (litter and dead wood; figure 2b). Ecosystems accumulated carbon in all scenarios but at different rates, with trajectories shaped primarily by climate change and to a lesser extent by land-use change. The highest and most consistent projected accumulation of ecosystem carbon occured under the combination of low radiative forcing and low land use change, yielding a ~15% increase in ecosystem carbon to an average of 363 Tg by 2100 (figure 2a). In contrast, high radiative forcing and high land use change resulted in the lowest ecosystem carbon gain, reaching a peak of ~332 Tg in 2063 and a decline to 327 Tg in 2100, resulting in a net increase of only 3% by the end of the simulation period (figure 2a). Ecosystem carbon accumulation was driven exclusively by increasing soil organic carbon across all four scenarios, all other stocks declined over time (figure 2b).

@@

### *3.2 Ecosystem carbon balance*
Net biome productivity (NBP) averaged approximately 0.6 Tg C y^-1^ at the start of the simulation period and declined over time in all four scenarios (figure 4). On average, terrestrial ecosystems of the seven main Hawaiian Islands collectively acted as a net carbon sink throughout the simulation period under the RCP 4.5 radiative forcing scenario, but carbon sink strength was ~40% lower in the high land-use scenario compared to the low land-use scenario by the end of the simulation period (figure 4). In contrast, ecosystems of the Hawaiian Islands acted as a net carbon source to the atmosphere toward the latter half of the simulation period under RCP 8.5, with the transition from sink to source occuring 15 years earlier on average in the high land-use scenario than in the low land-use scenario (figure 4). The high land-use scenario under RCP 8.5 represented a ~40% larger net source of carbon to the atmosphere by the year 2100 than the low-land use scenario under the same radiative forcing. Over the entire simulation period, both global emissions reductions and local avoided land conversion resulted in substantial increases in cumulative NBP (figure 5). However, switching from RCP 8.5 to RCP 4.5 increased cumulative NBP in the Hawaiian Islands more than twice as much as reducing emissions from local land-use change and wildfire disturbance (figure 5). Switching from RCP 8.5 to RCP 4.5 under the low land-use scenario yielded the greatest cumulative increase in NBP, resulting in a median gain of 26.5 Tg of carbon over the entire 90-year simulation period.

@@

### *3.3 CO~2~ fertilization effect*
Projected estimates of both total ecosystem carbon storage and ecosystem carbon balance were highly sensitive to differing rates of a CFE on plant productivity. Under the high radiative forcing (RCP 8.5) and high land-use scenario, the inclusion of a CFE ranging from 5-15% led to ~33-98 Tg of additional carbon storage in ecosystems by the end of the century, a ~10-30% increase (figure 6a). Compared to the reference scenario (0% CFE), a 5% CFE was sufficient to transform Hawaiian Island ecosystems from a net carbon source to the atmosphere during the latter half of the 21st century (figure 4b) to a net carbon sink for the entire simulation period (figure 6b), completely offsetting all other carbon losses induced by high radiative forcing and high land use. Net carbon sink strength was further enhanced at higher CFE rates, with NBP increasing by an average of 0.07 Tg C y^-1^ for each 1% increase in CFE (figure 6b). When compared to other scenarios, applying a 5% CFE to the high radiative forcing and high land-use scenario resulted in a mean annual NBP of 0.46 ± 0.3 Tg C y^-1^, roughly equivalent to mean annual NBP in the low radiative forcing and low land-use scenario with no CFE (0.52 ± 0.12). A 15% CFE applied to the high radiative forcing and high land-use scenario resulted in a mean annual NBP of 1.18 ± 0.29 Tg C y^-1^, more than double that of the low radiative forcing and low land-use scenario with no CFE.

@@

## 4. Discussion

We estimated that terrestrial ecosystems of the Hawaiian Islands have been a consistent net sink for atmospheric carbon over the last decade (figure 4). For the time period 2011-2019, net biome productivity (NBP) averaged 0.64 TgC y^-1^ and ranged from 0.46 to 0.88 TgC y^-1^ across all scenarios. Based on this mean annual NBP estimate, Hawaiian terrestrial ecosystems offset approximately 13% of 2015 statewide CO~2~ emissions from energy production and transportation (5.04 TgC), the State of Hawaiʻi's largest source of greenhouse gas emissions [@hawaii_inventory_2019]. Future projections indicate Hawaiian terrestrial ecosystems will continue to be a net sink for atmospheric carbon if global CO~2~ emissions peak around 2040 and then decline (RCP 4.5), and that carbon sink strength can be further enhanced by reducing the intensity and extent of future land use change. If, however, global CO~2~ emissions continue to rise throughout the 21st century (RCP 8.5), our projections indicate Hawaiian ecosystems will transition from a net sink to a net source of CO~2~ to the atmosphere, with high levels of land-use change accelerating this transition and exacerbating net carbon loss. Our model results also indicate that projections of ecosystem carbon balance are highly sensitive to the introduction of a CFE. Even a 5% increase in NPP for every 100ppm increase in atmospheric CO~2~ was sufficient to completely offset all other carbon losses induced by the high radiative forcing and high land use scenario, maintaining Hawaiian Island ecosystems as a net carbon sink for the entire simulation period instead of transitioning to a net carbon source by mid-century. Reconciling the high uncertainty surrounding the response of net photosynthesis to rising atmospheric CO~2~ is essential to more realistically constrain model projections of ecosystem carbon balance.

### *4.1 Drivers of ecosystem carbon balance*

### *4.2 Comparison to other studies*
There are few estimates of contemporary ecosystem carbon balance for the main Hawaiian Islands, and even fewer model projections of future ecosystem carbon balance in response to climate and land use change.  Our mean annual NBP estimate of 0.64 TgC y^-1^ for the period 2011-2019 agrees well with a recent State of Hawaiʻi Greenhouse Gas Inventory report, which estimated an annual net carbon sink of 0.66 TgC in 2015 from agriculture, forestry, and other land uses [@hawaii_inventory_2019].

### *4.3 Limitations of this study*

## 5. Conclusion

### 2021-02-16 Manuscript/HI_Model_supp.Rmd
updated manuscript files, added MOD17 NPP data and R code, updated README file
updated Fire R scripts and manuscript files

---
title: "Supplemental material: Ecosystem carbon balance in the Hawaiian Islands under different scenarios of future climate and land use change"
bibliography: HI_Model.bib
csl: iop.csl
geometry: margin=1in
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
output: 
  pdf_document: 
    latex_engine: xelatex
mainfont: Times New Roman
fontsize: 12pt
urlcolor: blue

---

@@

>>>>

---
title: "Supplemental material: Ecosystem carbon balance in the Hawaiian Islands under different scenarios of future climate and land use change"
bibliography: HI_Model.bib
csl: iop.csl
geometry: margin=1in
header-includes:
   - \usepackage{lineno}
   - \linenumbers
   - \usepackage{setspace}
   - \doublespacing
output: 
  pdf_document: 
    latex_engine: xelatex
mainfont: Times New Roman
fontsize: 12pt
urlcolor: blue
nocite: |
  @jacobi_map_2017
---

@@

## Moisture Zones

State Classes for Agriculture, Forest, Grassland, Shrubland, and Tree Plantations were stratified into three Moisture Zones - Dry, Mesic, and Wet (Figure S1). These three zones were based on a moisture availability index (MAI), calculated as mean annual precipitation (MAP) minus potential evapotranspiration [PET; @price_mz_2012]. Areas where the MAI was less than zero, i.e. where PET exceeded MAP, were classified in the Dry Moisture Zone. Areas with MAI values between zero and 1,661 were classified in the Mesic Moisture Zone. The MAI value of 1,661 is roughly equivalent to areas at 1,000m elevation that receive 2,500 mm of annual rainfall [@price_mz_2012]. Areas with MAI values that exceeded 1,661 were classified in the Wet Mesic Zone.

```{r fig.align = 'center', out.width = "99%", fig.cap = "Moisture zones of the seven main Hawaiian Islands, adapted from Jacobi et al (2017)."}
knitr::include_graphics("./fig_images/figS1_MZ.png")
```

## Wildland Fire

This study used a new spatial database of wildland fire perimeters in the Hawaiian Islands for the years 1999-2019 compiled using data from the Monitoring Trends in Burn Severity (MTBS) program, the Hawaiʻi Wildfire Management Organization, the U.S. National Park Service, the Oʻahu Army Natural Resources Program, and independent remote sensing analyses using publicly available LandSat data. This new wildland fire spatial database was used to calculate annual area burned (Figure S2), wildland fire probabilities, and wildland fire size distributions.

```{r fig.align = 'center', out.width = "99%", fig.cap = "Annual area burned by wildland fire in the State of Hawaiʻi from 1999-2019, summed by year across the seven main Hawaiian Islands (A) and within each of the four largest islands (B). The dashed horizontal line in (A) represents the median area burned from 1999-2019."}
knitr::include_graphics("./fig_images/figS2_AnnualAreaBurned.png")
```

## Climate

Spatially explicit mean annual data on contemporary temperature and rainfall for the main Hawaiian Islands at 250-m resolution (Figure S3) are from @giambelluca_online_2013 and @giambelluca_evapotranspiration_2014. Spatially explicit projections of changes in annual temperature and rainfall by the year 2100 were developed using statistically downscaled CMIP5 climate projections for the Hawaiian Islands from @timm_statistical_2015 and @timm_future_2017 under RCPs 4.5 and 8.5 (Figure S5).

```{r fig.align = 'center', out.width = "99%", fig.cap = "Mean annual temperature (top panel) and mean annual rainfall (bottom panel) for the seven main Hawaiian Islands."}
knitr::include_graphics("./fig_images/figS3_CurrentClimate.png")
```

```{r fig.align = 'center', out.width = "99%", fig.cap = "Projected change in mean annual temperature (top panels) and mean annual rainfall (bottom panels) by end of century under RCPs 4.5 and 8.5 (top panels) based on statistical downscaling of an ensemble of climate models (CMIP5)."}
knitr::include_graphics("./fig_images/figS4_FutureClimate.png")
```

## Change in State Class Area

Projections of total land area covered by each State Class were summed by year across the seven main Hawaiian Islands from LUCAS model output data.

```{r fig.align = 'center', out.width = "99%", fig.cap = "Projections of total land area by year for each of five State Classes in the seven main Hawaiian Islands (Agriculture, Developed, Forest, Grassland, and Shrubland) under low and high land use change scenarios for the period 2010-2100. Solid lines represent the mean of 30 Monte Carlo realizations and shaded areas represent minimum and maximum Monte Carlo values."}
knitr::include_graphics("./fig_images/figS5_StateClassArea.png")
```

## References
